Hazelnut: A Bidirectionally Typed \\ Structure Editor Calculus

### 2016-11-15 formally.tex
dropped construct arg per reviewer comment / lack of symmetry
revised abstract and intro

The previous section introduced Hazelnut by example. In this section, we
systematically introduce the following structures:
\begin{itemize}[itemsep=0px,partopsep=2px,topsep=2px]
\item \textbf{H-types} and \textbf{H-expressions} (Sec. \ref{sec:holes}),
  which are types and expressions with {holes}. H-types classify
  H-expressions according to Hazelnut's \textbf{bidirectional static
    semantics}.

@@

Our overview below omits certain ``uninteresting'' details. The appendix
includes the complete collection of rules, in definitional order. These
rules, along with the proofs of all of the metatheorems discussed, have been
mechanized in Agda \cite{norell:thesis} (see Sec. \ref{sec:mech}.)

\subsection{H-types and H-expressions}\label{sec:holes}
Figure \ref{fig:hexp-syntax} defines the syntax of H-types, $\htau$, and
H-expressions, $\hexp$. Most forms correspond directly to those of the
simply typed lambda calculus (STLC) extended with a single base type,
$\tnum$, of numbers (cf. \cite{pfpl}.) The number expression corresponding
to the mathematical number $n$ is drawn $\hnum{n}$, and for simplicity, we
define only a single arithmetic operation, $\hadd{\hexp}{\hexp}$. The form
$\hexp : \htau$ is an explicit \emph{type ascription}.

In addition to these standard forms, \emph{type holes} and \emph{empty
  expression holes} are both drawn $\hehole$ and \emph{non-empty expression
  holes} are drawn $\hhole{\hexp}$. We do not need non-empty type holes
because every H-type is a valid classifier of H-expressions.

Types and expressions that contain no holes are \emph{complete types} and
\emph{complete expressions}, respectively. Formally, we can derive
$\hcomplete{\htau}$ when $\htau$ is complete, and $\hcomplete{\hexp}$ when
$\hexp$ is complete (see supplement.) We are not concerned here with
defining a dynamic semantics for Hazelnut, but the dynamics for complete
H-expressions would be entirely standard because complete expressions correspond  immediately to those of the typed lambda calculus with numbers (in Sec. \ref{sec:rw} we discuss
the multi-dimensional design space around the question of evaluating
incomplete H-expressions.)

\begin{figure}[t]
$\arraycolsep=4pt\begin{array}{lllllll}
\mathsf{HTyp} & \htau & ::= &
  \tarr{\htau}{\htau} ~\vert~
  \tnum ~\vert~
  \tehole\\
\mathsf{HExp} & \hexp & ::= &
  \hexp : \htau ~\vert~
  x ~\vert~
  \hlam{x}{\hexp} ~\vert~
  \hap{\hexp}{\hexp} ~\vert~
  \hnum{n} ~\vert~
  \hadd{\hexp}{\hexp} ~\vert~
  \hehole ~\vert~
  \hhole{\hexp}
\end{array}$
\caption{Syntax of H-types and H-expressions. Metavariable $x$ ranges over variables and $n$ ranges over numerals.}
\label{fig:hexp-syntax}
\end{figure}
\begin{figure}
\noindent\fbox{$\tcompat{\htau}{\htau'}$}~~\text{$\tau$ and $\tau'$ are consistent}
\begin{subequations}\label{eqns:consistency}
\begin{mathpar}
\inferrule{ }{
  \tcompat{\tehole}{\htau}
}
~~~~~
\inferrule{ }{
  \tcompat{\htau}{\tehole}
}
~~~~~
\inferrule{ }{
  \tcompat{\htau}{\htau}
}
~~~~~
\inferrule{
  \tcompat{\htau_1}{\htau_1'}\\
  \tcompat{\htau_2}{\htau_2'}
}{
  \tcompat{(\tarr{\htau_1}{\htau_2})}{(\tarr{\htau_1'}{\htau_2'})}
}~~~~\text{\hspace{-2px}(\ref*{eqns:consistency}a-d)}
\end{mathpar}
\end{subequations}
\fbox{$\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}$}~~\text{$\tau$ has matched arrow type $\tarr{\htau_1}{\htau_2}$}
\begin{subequations}
\begin{minipage}{0.5\linewidth}
\begin{equation}
\inferrule{ }{
  \arrmatch{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1}{\htau_2}}
}
\end{equation}
\end{minipage}
\begin{minipage}{0.48\linewidth}
\begin{equation}
\inferrule{ }{
  \arrmatch{\tehole}{\tarr{\tehole}{\tehole}}
}
\end{equation}
\end{minipage}
\end{subequations}
\caption{H-type consistency and matched arrow types.}
\label{fig:type-consistency}
\end{figure}

\begin{figure}
\fbox{$\hsyn{\hGamma}{\hexp}{\htau}$}~~\text{$\hexp$ synthesizes $\htau$}
\begin{subequations}
\begin{equation}\label{rule:syn-asc}
\inferrule{
  \hana{\hGamma}{\hexp}{\htau}
}{
  \hsyn{\hGamma}{\hexp : \htau}{\htau}
}
\end{equation}
\begin{equation}\label{rule:syn-var}
\inferrule{ }{
  \hsyn{\hGamma, x : \htau}{x}{\htau}
}
\end{equation}
\begin{equation}\label{rule:syn-ap}
\inferrule{
  \hsyn{\hGamma}{\hexp_1}{\htau}\\
  \arrmatch{\htau}{\tarr{\htau_2}{\htau'}}\\
  \hana{\hGamma}{\hexp_2}{\htau_2}
}{
  \hsyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\htau'}
}
\end{equation}
\begin{equation}\label{rule:syn-num}
\inferrule{ }{
  \hsyn{\hGamma}{\hnum{n}}{\tnum}
}
\end{equation}
\begin{equation}\label{rule:syn-plus}
\inferrule{
  \hana{\hGamma}{\hexp_1}{\tnum}\\
  \hana{\hGamma}{\hexp_2}{\tnum}
}{
  \hsyn{\hGamma}{\hadd{\hexp_1}{\hexp_2}}{\tnum}
}
\end{equation}
\begin{equation}\label{rule:syn-ehole}
\inferrule{ }{
  \hsyn{\hGamma}{\hehole}{\tehole}
}
\end{equation}
\begin{equation}\label{rule:syn-hole}
\inferrule{
  \hsyn{\hGamma}{\hexp}{\htau}
}{
  \hsyn{\hGamma}{\hhole{\hexp}}{\tehole}
}
\end{equation}
\end{subequations}
\fbox{$\hana{\hGamma}{\hexp}{\htau}$}~~\text{$\hexp$ analyzes against $\htau$}
\begin{subequations}
\begin{equation}\label{rule:ana-subsume}
\inferrule{
  \hsyn{\hGamma}{\hexp}{\htau'}\\
  \tcompat{\htau}{\htau'}
}{
  \hana{\hGamma}{\hexp}{\htau}
}
\end{equation}
\begin{equation}\label{rule:syn-lam}
\inferrule{
  \arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}\\
  \hana{\hGamma, x : \htau_1}{\hexp}{\htau_2}
}{
  \hana{\hGamma}{\hlam{x}{\hexp}}{\htau}
}
\end{equation}
\end{subequations}
\caption{Synthesis and analysis.}
\label{fig:ana-syn}
\end{figure}

Hazelnut's static semantics is organized as a \emph{bidirectional type
  system}
\cite{Pierce:2000:LTI:345099.345100,DBLP:conf/icfp/DaviesP00,DBLP:conf/tldi/ChlipalaPH05,bidi-tutorial}
around the two mutually defined judgements in Figure
\ref{fig:ana-syn}. Typing contexts, $\hGamma$, map each variable $x \in
\domof{\hGamma}$ to a hypothesis $x : \htau$ in the usual manner. We
identify the context up to exchange and contraction and adopt the standard identification
convention for formal structures that differ only up to alpha-renaming of bound variables. All judgements that mention a context obey weakening.

Derivations of the type analysis judgement, $\hana{\hGamma}{\hexp}{\htau}$,
establish that $\hexp$ can appear where an expression of type $\htau$ is
expected. Derivations of the type synthesis judgement,
$\hsyn{\hGamma}{\hexp}{\htau}$, infer a type from $\hexp$, which is
necessary in positions where an expected type is not available (e.g. at the
top level.) Algorithmically, the type is an ``input'' of the type analysis
judgement, but an ``output'' of the type synthesis judgement.  Making a
judgemental distinction between these two notions will be essential in our
action semantics (Sec. \ref{sec:actions}.)

\begin{subequations}\label{rules:syn-ana}
Type synthesis is stronger than type analysis in that if an expression is
able to synthesize a type, it can also be analyzed against that type, or
any other \emph{consistent} type, according to the \emph{subsumption rule},
Rule (\ref{rule:ana-subsume}).

@@

Rule (\ref{rule:syn-lam}) defines analysis for lambda abstractions,
$\hlam{x}{\hexp}$. There is no type synthesis rule that applies to this
form, so lambda abstractions can appear only in analytic position,
i.e. where an expected type is known.\footnote{It is possible to also
  define a ``half-annotated'' synthetic lambda form, $\lambda x{:}\tau.e$,
  but for simplicity, we leave it out \cite{DBLP:conf/tldi/ChlipalaPH05}.}
Rule (\ref{rule:syn-lam}) is not quite the standard rule, reproduced below:
\begin{equation*}
\inferrule{
  \hana{\hGamma, x : \htau_1}{\hexp}{\htau_2}
}{
  \hana{\hGamma}{\hlam{x}{\hexp}}{\tarr{\htau_1}{\htau_2}}
}
\end{equation*}
The problem is that this standard rule alone leaves us unable to analyze
lambda abstractions against the type hole, because the type hole is not
immediately of the form $\tarr{\htau_1}{\htau_2}$. There are two plausible
solutions to this problem. One solution would be to define a second rule
specifically for this case:
\begin{equation*}
\inferrule{
  \hana{\hGamma, x : \tehole}{\hexp}{\tehole}
}{
  \hana{\hGamma}{\hlam{x}{\hexp}}{\tehole}
}
\end{equation*}
Instead, we combine these two possible rules into a single rule through the
simple auxiliary \emph{matched arrow type} judgement,
$\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}$, defined in Figure
\ref{fig:type-consistency}. This judgement leaves arrow types alone and
assigns the type hole the matched arrow type $\tarr{\tehole}{\tehole}$. It
is easy to see that the two rules above are admissible by appeal to Rule
(\ref{rule:syn-lam}) and the matched arrow type judgement. Encouragingly,
this judgement also arises in the study of gradual type systems
\cite{DBLP:conf/popl/CiminiS16,DBLP:conf/popl/GarciaC15,DBLP:conf/popl/RastogiCH12}.

Rule (\ref{rule:syn-asc}) defines type synthesis of expressions of
ascription form, $\hexp : \htau$. This allows the user to explicitly state
a type for the ascribed expression to be analyzed against.

Rule (\ref{rule:syn-var}) is the standard rule for variables.

Rule (\ref{rule:syn-ap}) is again nearly the standard rule for function
application. It also makes use of the matched function type judgement to
combine what would otherwise need to be two rules for function application
-- one for when $e_1$ synthesizes an arrow type, and another for when $e_1$
synthesizes the type hole. Indeed, Siek and Taha needed two application
rules for the same fundamental reason \cite{Siek06a}. Later work on gradual
typing introduced this notion of type matching to resolve this redundancy.

@@

@@

Rule (\ref{rule:syn-ehole}) states that the empty expression hole
synthesizes the type hole.

A non-empty hole contains an H-expression that is ``under construction'',
as described in Sec. \ref{sec:example}. According to Rule
(\ref{rule:syn-hole}), this inner expression must synthesize some type,
but, like the empty hole, non-empty holes synthesize the type hole.

Given these rules, it is instructive to derive the following (by subsumption):
\[\hana{incr : \tarr{\tnum}{\tnum}}{\hhole{incr}}{\tnum}\]

\end{subequations}
\subsection{Z-types and Z-expressions}\label{sec:cursors}
\newcommand{\cvert}{{\,{\vert}\,}}
\begin{figure}[t]
\hspace{-3px}$\arraycolsep=2pt\begin{array}{lllllll}
\mathsf{ZTyp} & \ztau & ::= &
  \zwsel{\htau} \cvert
  \tarr{\ztau}{\htau} \cvert
  \tarr{\htau}{\ztau} \\
\mathsf{ZExp} & \zexp & ::= &
  \zwsel{\hexp} \cvert
  \zexp : \htau \cvert
  \hexp : \ztau \cvert
  \hlam{x}{\zexp} \cvert
  \hap{\zexp}{\hexp} \cvert
  \hap{\hexp}{\zexp} \cvert
  \hadd{\zexp}{\hexp} \cvert
  \hadd{\hexp}{\zexp} \cvert
  \hhole{\zexp}
\end{array}$
\caption{Syntax of Z-types and Z-expressions.}
\label{fig:zexp-syntax}
\end{figure}

Hazelnut's action semantics operates over Z-types, $\ztau$, and
Z-expressions, $\zexp$. Figure \ref{fig:zexp-syntax} defines the syntax of
Z-types and Z-expressions. A Z-type (resp. Z-expression) represents an
H-type (resp. H-expression) with a single superimposed \emph{cursor}.

@@

We write $\removeSel{\ztau}$ for the H-type constructed by erasing the
cursor from $\ztau$, which we refer to as the \emph{cursor erasure} of
$\ztau$. This straightforward metafunction is defined as follows:
\begin{align*}
\removeSel{(\zwsel{\htau})} & = \htau\\
\removeSel{(\tarr{\ztau}{\htau})} & = \tarr{\removeSel{\ztau}}{\htau}\\
\removeSel{(\tarr{\htau}{\ztau})} & = \tarr{\htau}{\removeSel{\ztau}}
\end{align*}

Similarly, we write $\removeSel{\zexp}$ for the H-expression constructed by
erasing the cursor from the Z-expression $\zexp$, i.e. the cursor erasure
of $\zexp$. The definition of this metafunction is analogous, so we omit it
for concision (see supplement.)

@@

In some Z-expressions, the cursor is in a type ascription, so we also need
a \emph{type action judgement}, pronounced ``performing $\alpha$ on $\ztau$
results in $\ztau'$'':
\[
\performTyp{\ztau}{\alpha}{\ztau'}
\]

@@

Before giving the rules for these judgements, let us state the key
metatheorem: \emph{sensibility}. This metatheorem deeply informs the design
of the rules, given starting in Sec. \ref{sec:action-subsumption}. Its
proof, which is mechanized in Agda, is by straightforward induction, so the
reader is encouraged to think about the relevant cases of these proofs as
we present the rules.
\begin{theorem}[Action Sensibility]
  \label{thrm:actsafe} Both of the following hold:
  \begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
  \item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and
    $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$ then
    $\hsyn{\hGamma}{\removeSel{\zexp'}}{\htau'}$.
  \item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
    $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}$ then
    $\hana{\hGamma}{\removeSel{\zexp'}}{\htau}$.
  \end{enumerate}
\end{theorem}
\noindent In other words, if an edit state (i.e. a Z-expression) is
statically meaningful, i.e. its cursor erasure is well-typed, then
performing an action on it leaves the resulting edit state statically
meaningful. In particular, the first clause of Theorem \ref{thrm:actsafe}
establishes that when an action is performed on an edit state whose cursor
erasure synthesizes an H-type, the result is an edit state whose cursor
erasure also synthesizes some (possibly different) H-type. The second
clause establishes that when an action is performed using the analytic
action judgement on an edit state whose cursor erasure analyzes against
some H-type, the result is a Z-expression whose cursor erasure also
analyzes against the same H-type.

\begin{figure}[t]
\hspace{-3px}$\arraycolsep=3pt\begin{array}{llcllll}
\mathsf{Action} & \alpha & ::= &
  \aMove{\delta} ~\vert~
  \aConstruct{\psi} ~\vert~
  \aDel ~\vert~
  \aFinish\\
\mathsf{Dir} & \delta & ::= &
  \dChildnm{n} ~\vert~
  \dParent\\
\mathsf{Shape} & \psi & ::= &
  \farr ~\vert~
  \fnum \\
& & \vert &
  \fasc ~\vert~
  \fvar{x} ~\vert~
  \flam{x} ~\vert~
  \fap ~\vert~
  \farg ~\vert~
  \fnumlit{n} ~\vert~
  \fplus\\
& & \vert &
  {\color{gray}\fnehole}
\end{array}$
\caption{Syntax of actions.}
\label{fig:action-syntax}
\end{figure}

\subsubsection{Type Inconsistency}
In some of the rules below, we will need to supplement our definition of
type consistency from Figure \ref{fig:type-consistency} with a definition
of \emph{type inconsistency}, written $\tincompat{\htau}{\htau'}$. One can
define this notion either directly as the metatheoretic negation of type
consistency, or as a separate inductively defined judgement. The supplement
does the latter. The key rule establishes that arrow types are inconsistent
with $\tnum$:
  \begin{equation*}
    \inferrule{ }{
      \tincompat{\tnum}{\tarr{\htau_1}{\htau_2}}
    }
  \end{equation*}
The mechanization proves that this judgemental definition of type
inconsistency is indeed the negation of type consistency.

@@

The rules above are numerous and fairly uninteresting. That makes them
quite hazardous -- we might make a mistake absent-mindedly. One check
against this is to prove that establishes that movement actions do not
change the cursor erasure, as in Theorem \ref{lemma:movement-erasure}.

>>>>

The previous section introduced Hazelnut by example. In this section, we
systematically define the following structures:
\begin{itemize}[itemsep=0px,partopsep=2px,topsep=2px]
\item \textbf{H-types} and \textbf{H-expressions} (Sec. \ref{sec:holes}),
  which are types and expressions with {holes}. H-types classify
  H-expressions according to Hazelnut's \textbf{bidirectional static
    semantics}.

@@

Our overview below omits certain ``uninteresting'' details. The supplement 
includes the complete collection of rules, in definitional order. These
rules, along with the proofs of all of the metatheorems discussed in this section (and several omitted auxiliary lemmas), have been
mechanized using the Agda proof assistant \cite{norell:thesis} (discussed in Sec. \ref{sec:mech}.)

\subsection{H-types and H-expressions}\label{sec:holes}
Figure \ref{fig:hexp-syntax} defines the syntax of H-types, $\htau$, and
H-expressions, $\hexp$. Most forms correspond directly to those of the
simply typed lambda calculus (STLC) extended with a single base type,
$\tnum$, of numbers (cf. \cite{pfpl}.) The number expression corresponding
to the mathematical number $n$ is drawn $\hnum{n}$, and for simplicity, we
define only a single arithmetic operation, $\hadd{\hexp}{\hexp}$. The form
$\hexp : \htau$ is an explicit \emph{type ascription}. In addition to these standard forms, \emph{type holes} and \emph{empty
  expression holes} are both drawn $\hehole$ and \emph{non-empty expression
  holes} are drawn $\hhole{\hexp}$. 
Types and expressions that contain no holes are \emph{complete types} and
\emph{complete expressions}, respectively. 
%We do not, therefore, need to separately define the dynamics of complete expressions here (see, e.g., Harper \cite{pfpl}.) We will not develop a dynamics for incomplete expressions (but see Sec. \ref{sec:rw} for a discussion of this intriguing idea.) 
%Formally, we can derive
%$\hcomplete{\htau}$ when $\htau$ is complete, and $\hcomplete{\hexp}$ when
%$\hexp$ is complete (see supplement.)

\begin{figure}[t]
$\arraycolsep=4pt\begin{array}{lllllll}
\mathsf{HTyp} & \htau & ::= &
  \tarr{\htau}{\htau} ~\vert~
  \tnum ~\vert~
  \tehole\\
\mathsf{HExp} & \hexp & ::= &
  x ~\vert~
  {\hlam{x}{\hexp}} ~\vert~
  \hap{\hexp}{\hexp} ~\vert~
  \hnum{n} ~\vert~
  \hadd{\hexp}{\hexp} ~\vert~
  \hexp : \htau ~\vert~
  \hehole ~\vert~
  \hhole{\hexp}
\end{array}$
\caption{Syntax of H-types and H-expressions. Metavariable $x$ ranges over variables and $n$ ranges over numerals.}
\label{fig:hexp-syntax}
\end{figure}

\begin{figure}
\fbox{$\hsyn{\hGamma}{\hexp}{\htau}$}~~\text{$\hexp$ synthesizes $\htau$}
\begin{subequations}
\begin{equation}\label{rule:syn-var}
\inferrule{ }{
  \hsyn{\hGamma, x : \htau}{x}{\htau}
}
\end{equation}
\begin{equation}\label{rule:syn-ap}
\inferrule{
  \hsyn{\hGamma}{\hexp_1}{\htau_1}\\
  \arrmatch{\htau_1}{\tarr{\htau_2}{\htau}}\\
  \hana{\hGamma}{\hexp_2}{\htau_2}
}{
  \hsyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\htau}
}
\end{equation}
\begin{equation}\label{rule:syn-num}
\inferrule{ }{
  \hsyn{\hGamma}{\hnum{n}}{\tnum}
}
\end{equation}
\begin{equation}\label{rule:syn-plus}
\inferrule{
  \hana{\hGamma}{\hexp_1}{\tnum}\\
  \hana{\hGamma}{\hexp_2}{\tnum}
}{
  \hsyn{\hGamma}{\hadd{\hexp_1}{\hexp_2}}{\tnum}
}
\end{equation}
\begin{equation}\label{rule:syn-asc}
\inferrule{
  \hana{\hGamma}{\hexp}{\htau}
}{
  \hsyn{\hGamma}{\hexp : \htau}{\htau}
}
\end{equation}
\begin{equation}\label{rule:syn-ehole}
\inferrule{ }{
  \hsyn{\hGamma}{\hehole}{\tehole}
}
\end{equation}
\begin{equation}\label{rule:syn-hole}
\inferrule{
  \hsyn{\hGamma}{\hexp}{\htau}
}{
  \hsyn{\hGamma}{\hhole{\hexp}}{\tehole}
}
\end{equation}
\end{subequations}
\fbox{$\hana{\hGamma}{\hexp}{\htau}$}~~\text{$\hexp$ analyzes against $\htau$}
\begin{subequations}
\begin{equation}\label{rule:syn-lam}
\inferrule{
  \arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}\\
  \hana{\hGamma, x : \htau_1}{\hexp}{\htau_2}
}{
  \hana{\hGamma}{\hlam{x}{\hexp}}{\htau}
}
\end{equation}
\begin{equation}\label{rule:ana-subsume}
\inferrule{
  \hsyn{\hGamma}{\hexp}{\htau'}\\
  \tcompat{\htau}{\htau'}
}{
  \hana{\hGamma}{\hexp}{\htau}
}
\end{equation}
\end{subequations}
\caption{H-type synthesis and analysis.}
\label{fig:ana-syn}
\end{figure}
\begin{figure}
\vspace{-1px}\noindent\fbox{$\tcompat{\htau}{\htau'}$}~~\text{$\htau$ and $\htau'$ are consistent}\vspace{-4px}
\begin{subequations}\label{eqns:consistency}
\begin{mathpar}
\inferrule{ }{
  \tcompat{\tehole}{\htau}
}
~~~~~
\inferrule{ }{
  \tcompat{\htau}{\tehole}
}
~~~~~
\inferrule{ }{
  \tcompat{\htau}{\htau}
}
~~~~~
\inferrule{
  \tcompat{\htau_1}{\htau_1'}\\
  \tcompat{\htau_2}{\htau_2'}
}{
  \tcompat{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1'}{\htau_2'}}
}~~~~\text{\hspace{-2px}(\ref*{eqns:consistency}a-d)}
\end{mathpar}
\end{subequations}
\noindent\fbox{$\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}$}~~\text{$\htau$ has matched arrow type $\tarr{\htau_1}{\htau_2}$}\vspace{-4px}

\begin{subequations}
\begin{minipage}{0.43\linewidth}
\begin{equation}
\inferrule{ }{
  \arrmatch{\tehole}{\tarr{\tehole}{\tehole}}
}
\end{equation}
\end{minipage}
\begin{minipage}{0.55\linewidth}
\begin{equation}
\inferrule{ }{
  \arrmatch{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1}{\htau_2}}
}
\end{equation}
\end{minipage}
\end{subequations}
\caption{H-type consistency and matched arrow types.}
\label{fig:type-consistency}
\end{figure}

Hazelnut's static semantics is organized as a \emph{bidirectional type
  system}
\cite{Pierce:2000:LTI:345099.345100,DBLP:conf/icfp/DaviesP00,DBLP:conf/tldi/ChlipalaPH05,bidi-tutorial}
around the two mutually defined judgements in Figure
\ref{fig:ana-syn}. Derivations of the type analysis judgement, $\hana{\hGamma}{\hexp}{\htau}$,
establish that $\hexp$ can appear where an expression of type $\htau$ is
expected. Derivations of the type synthesis judgement,
$\hsyn{\hGamma}{\hexp}{\htau}$, synthesize (a.k.a. \emph{locally infer} \cite{Pierce:2000:LTI:345099.345100}) a type from $\hexp$. Type synthesis is
necessary in positions where an expected type is not available (e.g. at the
top level.) Algorithmically, the type is an ``input'' of the type analysis
judgement, but an ``output'' of the type synthesis judgement.  Making a
judgemental distinction between these two notions will be essential in our
action semantics (Sec. \ref{sec:actions}.)

If an expression is 
able to synthesize a type, it can also be analyzed against that type, or
any other \emph{consistent} type, according to the \emph{subsumption rule},
Rule (\ref{rule:ana-subsume}).% Type analysis of every form other than the lambda form goes through subsumption.

@@

Typing contexts, $\hGamma$, map each variable $x \in
\domof{\hGamma}$ to an hypothesis $x : \htau$. We
identify contexts up to exchange and contraction and adopt the standard identification
convention for structures that differ only up to alpha-renaming of bound variables. All hypothetical judgements obey a standard weakening lemma. Rule (\ref{rule:syn-var}) establishes that variable expressions synthesize the hypothesized H-type, in the standard manner.

Rule (\ref{rule:syn-lam}) defines analysis for lambda abstractions,
$\hlam{x}{\hexp}$. There is no type synthesis rule that applies to this
form, so lambda abstractions can appear only in analytic position,
i.e. where an expected type is known.\footnote{It is possible to also
  define a ``half-annotated'' synthetic lambda form, $\lambda x{:}\tau.e$,
  but for simplicity, we leave it out \cite{DBLP:conf/tldi/ChlipalaPH05}.}
Rule (\ref{rule:syn-lam}) is not quite the standard rule, as reproduced below:
\begin{equation*}
\inferrule{
  \hana{\hGamma, x : \htau_1}{\hexp}{\htau_2}
}{
  \hana{\hGamma}{\hlam{x}{\hexp}}{\tarr{\htau_1}{\htau_2}}
}
\end{equation*}
The problem is that this standard rule alone leaves us unable to analyze
lambda abstractions against the type hole, because the type hole is not
immediately of the form $\tarr{\htau_1}{\htau_2}$. There are two plausible
solutions to this problem. One solution would be to define a second rule
specifically for this case:
\begin{equation*}
\inferrule{
  \hana{\hGamma, x : \tehole}{\hexp}{\tehole}
}{
  \hana{\hGamma}{\hlam{x}{\hexp}}{\tehole}
}
\end{equation*}
Instead, we combine these two possible rules into a single rule through the
simple auxiliary \emph{matched arrow type} judgement,
$\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}$, defined in Figure
\ref{fig:type-consistency}. This judgement leaves arrow types alone and
assigns the type hole the matched arrow type $\tarr{\tehole}{\tehole}$. It
is easy to see that the two rules above are admissible by appeal to Rule
(\ref{rule:syn-lam}) and the matched arrow type judgement.  Encouragingly,
the matched arrow type judgement also arises in gradual type systems
\cite{DBLP:conf/popl/CiminiS16,DBLP:conf/popl/GarciaC15,DBLP:conf/popl/RastogiCH12}.

Rule (\ref{rule:syn-ap}) is again nearly the standard rule for function
application. It also makes use of the matched function type judgement to
combine what would otherwise need to be two rules for function application
-- one for when $e_1$ synthesizes an arrow type, and another for when $e_1$
synthesizes $\tehole$.

@@

Rule (\ref{rule:syn-asc}) defines type synthesis of expressions of
ascription form, $\hexp : \htau$. This allows the programmer to explicitly state
a type for the ascribed expression to be analyzed against.

@@

Rule (\ref{rule:syn-ehole}) states that the empty expression hole
synthesizes the type hole. Non-empty holes, which contain an H-expression that is ``under construction'' 
as described in Sec. \ref{sec:example}, also synthesize the hole type. According to Rule
(\ref{rule:syn-hole}), the enveloped expression must synthesize some (arbitrary) type. (We do not need non-empty type holes because every H-type is a valid classifier of H-expressions.)

Because the hole type is consistent with every type, expression holes can be analyzed against any type by subsumption. For example, it is instructive to derive the following:
\[\hana{incr : \tarr{\tnum}{\tnum}}{\hhole{incr}}{\tnum}\]

\subsection{Z-types and Z-expressions}\label{sec:cursors}
\newcommand{\cvert}{{\,{\vert}\,}}
\begin{figure}[t]
$\arraycolsep=2pt\begin{array}{lllllll}
\mathsf{ZTyp} & \ztau & ::= &
  \zwsel{\htau} \cvert
  \tarr{\ztau}{\htau} \cvert
  \tarr{\htau}{\ztau} \\
\mathsf{ZExp} & \zexp & ::= &
  \zwsel{\hexp} \cvert
  \hlam{x}{\zexp} \cvert
  \hap{\zexp}{\hexp} \cvert
  \hap{\hexp}{\zexp} \cvert
  \hadd{\zexp}{\hexp} \cvert
  \hadd{\hexp}{\zexp} \\
& & \cvert & 
    \zexp : \htau \cvert
  \hexp : \ztau \cvert
  \hhole{\zexp}
\end{array}$
\caption{Syntax of Z-types and Z-expressions.}
\label{fig:zexp-syntax}
\end{figure}

Figure \ref{fig:zexp-syntax} defines the syntax of
Z-types, $\ztau$, and Z-expressions, $\zexp$. A Z-type (resp. Z-expression) represents an
H-type (resp. H-expression) with a single superimposed \emph{cursor}.

@@

We write $\removeSel{\ztau}$ for the H-type constructed by erasing the
cursor from $\ztau$, which we refer to as the \emph{cursor erasure} of
$\ztau$. This straightforward metafunction is defined as follows:
\begin{align*}
\removeSel{\zwsel{\htau}} & = \htau\\
\removeSel{\tarr{\ztau}{\htau}} & = \tarr{\removeSel{\ztau}}{\htau}\\
\removeSel{\tarr{\htau}{\ztau}} & = \tarr{\htau}{\removeSel{\ztau}}
\end{align*}

Similarly, we write $\removeSel{\zexp}$ for cursor erasure of $\zexp$. The definition of this metafunction is analogous, so we omit it
for concision.

This zipper structure is not the only way to model a cursor, though we have found it to be the most elegant for our present purposes. Another plausible strategy would be to formalize the notion of a relative path into an H-expression. This would then require defining the notion of consistency between a relative path and an H-expression, so we avoid it.

@@

In some Z-expressions, the cursor is in a type ascription, so we also need
a \emph{type action judgement}:
\[
\performTyp{\ztau}{\alpha}{\ztau'}
\]

@@

These judgements are governed by a critical 
metatheorem, \emph{action sensibility} (or simply \emph{sensibility}): 
\begin{theorem}[Action Sensibility]
  \label{thrm:actsafe} ~
  \begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
  \item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and
    $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$ then
    $\hsyn{\hGamma}{\removeSel{\zexp'}}{\htau'}$.
  \item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
    $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}$ then
    $\hana{\hGamma}{\removeSel{\zexp'}}{\htau}$.
  \end{enumerate}
\end{theorem}
\noindent In other words, if a Z-expression is
statically meaningful, i.e. its cursor erasure is well-typed, then
performing an action on it leaves the resulting Z-expression statically
meaningful. More specifically, the first clause of Theorem \ref{thrm:actsafe}
establishes that when an action is performed on a Z-expression whose cursor
erasure synthesizes an H-type, the result is a Z-expression whose cursor
erasure also synthesizes some (possibly different) H-type. The second
clause establishes that when an action is performed using the analytic
action judgement on an edit state whose cursor erasure analyzes against
some H-type, the result is a Z-expression whose cursor erasure also
analyzes against the same H-type.

This metatheorem deeply informs the design
of the rules, given starting in Sec. \ref{sec:action-subsumption}. Its
proof is by straightforward induction, so the
reader is encouraged to think about the relevant proof case when considering each action rule below.

No sensibility theorem is needed for the type action judgement because every syntactically well-formed type is meaningful in Hazelnut. (Adding type variables to the language would require defining both a type-level sensibility theorem and 
type-level non-empty holes.)
\begin{figure}[t]
\hspace{-3px}$\arraycolsep=3pt\begin{array}{llcllll}
\mathsf{Action} & \alpha & ::= &
  \aMove{\delta} ~\vert~
  \aConstruct{\psi} ~\vert~
  \aDel ~\vert~
  \aFinish\\
\mathsf{Dir} & \delta & ::= &
  \dChildnm{n} ~\vert~
  \dParent\\
\mathsf{Shape} & \psi & ::= &
  \farr ~\vert~
  \fnum \\
& & \vert &
  \fasc ~\vert~
  \fvar{x} ~\vert~
  \flam{x} ~\vert~
  \fap ~\vert~
  % \farg ~\vert~
  \fnumlit{n} ~\vert~
  \fplus\\
& & \vert &
  {\color{gray}\fnehole}
\end{array}$
\caption{Syntax of actions.}
\label{fig:action-syntax}
\end{figure}

\subsubsection{Type Inconsistency}
In some of the rules below, we will need to supplement our definition of
type consistency from Figure \ref{fig:type-consistency} with a definition
of \emph{type inconsistency}, written $\tincompat{\htau}{\htau'}$. One can
define this notion either directly as the constructive negation of type
consistency, or as a separate inductively defined judgement with the following
key rule, which establishes that arrow types are inconsistent
with $\tnum$:
  \begin{equation*}
    \inferrule{ }{
      \tincompat{\tnum}{\tarr{\htau_1}{\htau_2}}
    }
  \end{equation*}
The mechanization proves that the judgemental definition of type
inconsistency is indeed the negation of type consistency.

@@

The rules above are numerous and fairly uninteresting. That makes them
quite hazardous -- we might make a typo or forget a rule absent-mindedly. One check
against this is to establish that movement actions do not
change the cursor erasure, as in Theorem \ref{lemma:movement-erasure}.
